---
title: "Gathering insights about COVID19 Worldwide Testing using Linear Mixed Models"
subtitle: "An analysis of linear mixed model algorithms and it's applications"

author: "Autumn Clark and Alex Ottley (Advisor: Dr.Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

[Literature Review](litreview.html){target="_blank"} |
[Slides](slides.html){target="_blank"}


## Introduction

Understanding complex, hierarchical data has become increasingly important across disciplines such as epidemiology, cardiology, psychology, and public health. Among the most powerful and flexible tools for analyzing such data is the Linear Mixed Model (LMM). LMMs extend traditional linear regression by incorporating both fixed effects, which account for population-level trends, and random effects, which capture variability across individual units or clusters. This structure enables LMMs to model correlated data—such as repeated measures from the same subject or observations nested within groups—without violating the assumption of independence required by simpler models.

The usefulness of LMMs is particularly evident in medical and health research, where individual variability is often substantial. For instance, [@Goncalves2005] highlighted the superior performance of LMMs over ANOVA in evaluating heart rate and coronary-perfused pressure data in transgenic mice. Their work demonstrated how LMMs provide a more sensitive and robust framework for assessing experimental effects in biomedical studies, particularly when data are unbalanced or collected over time.

In cardiovascular research, sleep disturbances have been identified as significant contributors to abnormal nocturnal blood pressure (BP) profiles—factors that increase the risk of adverse cardiac events. [@Matsumoto2019] employed LMMs in their investigation of sleep efficiency and nocturnal BP using multi-day observational data. Their use of random effects allowed them to control for between-subject variability while modeling day-to-day BP fluctuations, resulting in more accurate estimates of the associations. The study supported previous findings that lower sleep efficiency is linked to higher nighttime BP levels and diminished nocturnal BP dips [@Kario2001] [@Mano2021] 

The advantages of LMMs extend to the modeling of 24-hour physiological processes. [@EdwardsandSimpson2014] utilized orthonormal polynomials within LMMs to analyze 24-hour ambulatory blood pressure monitoring data. Their approach enhanced both the accuracy and interpretability of BP trajectories over time, providing clinicians with detailed insights into circadian rhythm variations in patient data. LMMs allowed for complex modeling without overfitting, an essential property when working with high-frequency longitudinal health data.

Moreover, LMMs have proven especially valuable in large-scale clinical outcomes research. [@Wang2022] showcased the strength of LMMs using a dataset from over 600 patients who underwent pulmonary valve replacement surgery. With more than 5,000 echocardiographic measurements over multiple years, the study leveraged LMMs to model both fixed effects (e.g., treatment procedures, age, sex) and random effects (e.g., patient-specific recovery trajectories). LMMs allowed the researchers to handle missing data, unbalanced measurement schedules, and inter-patient heterogeneity more effectively than traditional linear models or survival analysis.

Together, these studies establish LMMs as a critical tool in longitudinal and repeated-measures research, offering flexibility, accuracy, and interpretability. This paper builds on that foundation by applying LMMs to the COVID-19 Worldwide Testing dataset, which contains country-level, time-indexed data on daily testing and positivity rates. While the subject differs from cardiovascular data, the analytical challenges—such as repeated observations, between-country variability, and time trends—are strikingly similar. The goal is to illustrate how LMMs can uncover meaningful patterns in large public health datasets, ultimately guiding more informed decision-making in times of global crisis. Later sections will also briefly reflect on how these models might be applied in hypothetical clinical settings, such as circadian blood pressure analysis.


## 2 Methods

Linear Mixed Models (LMMs) are a powerful statistical approach for analyzing data with both fixed and random effects, commonly used when observations are correlated or grouped (i.e., repeated measures, hierarchical data). 

The selected dataset for this project represents repeated measures data, which is data that includes observations collected from the same variable (dependent variable) multiple times, either over time or under different conditions, allowing for analysis of changes or effects within those subjects.

The goal of this analysis is to present how COVID-19 testing relates to the number of positive cases across different countries. Although we expect a general relationship, more tests leading to more positive cases, the strength of that relationship likely varies from country to country due to differences in testing strategies, population size, healthcare infrastructure, and outbreak severity.
To capture this country level variability, we apply a Linear Mixed Model framework, which allows us to analyze both overall trends and country-specific deviations. It is also important to recognize why we are using a Linear Mixed Model to answer this question. A traditional linear regression would present a single slope and intercept for all countries, which ignores the natural variation between regions. A LMM allows each country to have its own intercept or baseline number of positive cases. An LMM also allows each country to have its own slope which is presented as the effect of testing on positivity.

## 2.1 Understanding Linear Mixed Models

A Linear Mixed Model (LMM) is a flexible extension of the linear regression model that incorporates both fixed and random effects. Fixed effects represent population-level parameters that are constant across all groups (e.g., the average impact of daily testing across all countries). Random effects capture variation across clusters (e.g., countries), allowing each group to have its own intercept and/or slope. This is particularly useful for hierarchical, repeated-measures, or panel data, where observations within the same group are not independent.

LMMs are ideal for modeling data where residual errors are correlated within groups but independent across groups, which violates the assumptions of ordinary least squares (OLS) regression. They also allow analysts to partition variability into within-group and between-group components, improving both inference accuracy and generalizability.

## 2.2 Model Structure and Notation

The general form of a linear mixed model is:

$$
y_{ij} = \beta_0 + \beta_1 x_{ij} + b_{0j} + b_{1j}x_{ij} + \varepsilon_{ij}
$$
Where,

- $y_{ij}$: response variable for observation in $i$ in group $j$

- $x_{ij}$: predictor (e.g., scaled daily tests)

- $\beta_0,\beta_1$: fixed intercept and slope (population-level)

- $b_{0j}, b_{1j}$: random intercept and slope for group $j$ (e.g., country)

- $\varepsilon_{ij}$: residual error, assumed to be normally distributed: $\varepsilon_{ij} \sim N (0, \sigma^2)$

In matrix form, the LMM can be represented as:
$$
y = X\beta + Zb + \varepsilon
$$
Where,

- $y$: vector of outcomes

- $X$: fixed-effects design matrix

- $\beta$: vector of fixed effect coefficients

- $Z$: random-effects design matrix

- $b$: vector of random effects, assumed $\sim N (0, G)$

- $\varepsilon$: residual errors, assumed $\sim N(0, R)$

This framework enables the estimation of both overall trends (fixed effects) and group-specific deviations (random effects), making linear mixed models ideal for handling variability and correlation structures across clusters.


## 2.3 Application to COVID-19 Data

In this study, we employed a linear mixed model to examine the relationship between the number of daily COVID-19 positive cases reported across various countries over time. The dataset consists of repeated daily observations for each country, which resultws in multilevel structure where each country is considered a separate group with its own data points over time. This hierarchical structure gives the opportunity to account for both the overall trends in the data and the variability across different countries.

The model was designed to account for both the fixed effects of the number of daily tests conducted, which can influence case detection, and the random effects that capture country-specific deviations in testing practices and positivity rates. The model allows the both the intercept, baseline positivity, and the slope, the effect of testing on reported cases, to vary by country. This is incredibly important because different countries may have varying baseline rates of COVID-19 positivity and differing testing strategies, which may influence the number of positive cases detected.

### Key components of the model:

- Response variable: *daily_positive* — the number of newly confirmed cases per day
- Fixed effect predictor: *daily_tested_scaled* — the number of daily tests conducted (scaled by 1,000)
- Random effects: Intercept and slope for each *Country_Region*, allowing both baseline positivity and test effectiveness to vary by country

The model formula in R was specified as:

*lmer(daily_positive ~ daily_tested_scaled + (daily_tested_scaled | Country_Region), data = df_clean*)*

This structure enables the model to capture both the overall trend of case detection as influenced by the number of tests conducted, while simultaneously accounting for the heterogeneity in testing practices and baseline positivity across countries. By allowing these country-level variations, the model provides a more accurate understanding of how COVID-19 testing impacted the reported cases in different settings.

## 2.4 Justification

Linear mixed models were chosen due to their ability to model nested data, adjust for within-group correlation, and incorporate individual variation in trends. These models also handle unbalanced data (i.e., unequal observations per country) and missingness better than traditional methods like repeated-measures ANOVA or fixed-effects OLS regression.

## 3 Analysis and Results


### 3.1 Data and Visualization

### Packages 

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
library(lme4)
library(nlme)
library(dplyr)
library(ggplot2)
library(readr)

```

### Data Import
```{r, warning=FALSE, echo=TRUE}
# Load Data
library(readr)
COVID19 <- read_csv("tested_worldwide.csv")
kable(head(COVID19))
df <- COVID19
  
```

Data Source: **Kaggle**

[Covid19 Worldwide Testing](https://www.kaggle.com/datasets/lin0li/covid19testing/data){target="_blank"}

**Description:** This dataset, which has the number of tests conducted over time, is important to help making sense of daily reported cases and understanding how COVID-19 is truly spreading in each country.
The world generally uses daily reported positive cases to understand how COVID-19 is spreading in different countries. However, given that potentially over 50% of the cases are asymptomatic, these populations are much less likely to seek testing and get tested. Therefore, the reported cases are highly likely to be only a portion of the population who are truly infected by COVID-19, and the number of reported cases has a strong relation to the number of tests conducted each day.

For example, imagine a country not doing much testing might be reporting low number of cases, but it doesn't mean COVID-19 is not spreading in that country. To help making sense of daily reported cases and understanding how COVID-19 is truly spreading in each country, it's crucial to understand how many tests are conducted each day.

##### **Dataset Columns:**

- **Date**: an identifier for each individual
- **Country_Region**: gender of the individual (male or female)
- **Province_State**: age of the individual in years
- **Positive**: cumulative number of positive cases reported
- **Active**: number of active cases on that day
- **Hospitalized**: cumulative number of hospitalized cases reported
- **HospitalizedCurr**: number of actively hospitalized cases on that day
- **Recovered**: cumulative number of recovered cases reported
- **Death**: cumulative number of deaths reported
- **Total_tested**: cumulative number of tests conducted
- **Daily_tested**: number of tests conducted on the day; if daily data is unavailable, daily tested is averaged across of days in
- **Daily_positive**: number of positives cases reported on the day; if daily data is unavailable, daily positive is averaged across number of days in

### 3.2 Exploratory Data Analysis (Clean Data)
```{r}

df <- COVID19
#Convert Date to Date format
df$Date <- as.Date(df$Date)

#Remove rows with missing values in relevant columns
df_clean <- df %>%
  filter(!is.na(daily_positive), !is.na(daily_tested), !is.na(Country_Region))

#Rescale daily_tested column to improve interpretation and convergence
df_clean$daily_tested_scaled <- df_clean$daily_tested / 1000

```

### Model

```{r}
#Fit LMM with random intercept and slope
model_refined <- lmer(daily_positive ~ daily_tested_scaled + (daily_tested_scaled | Country_Region), data = df_clean)

head(df_clean)

#Model Summary
summary(model_refined)

```

### 3.3 Model Specification
We modeled the number of daily positive COVID-19 cases (daily_positive) as a function of the number of daily tests conducted (daily_tested_scaled). We fit a Linear Mixed Model (LMM) with a random intercept and random slope for each country:

- Response variable: *daily_positive*
- Fixed effect predictor: *daily_tested_scaled* (daily tests per 1,000)
- Random effects: intercepts and slopes grouped by *Country_Region*

This model allows each country to have its own baseline number of positive cases and its own relationship between testing volume and positivity.

### 3.4 Fixed Effects Interpretation
```{r}
#Extract and print key fixed effects
fixef(model_refined) #estimates for population-level intercept and slope
```


The fixed effect for *daily_tested_scaled* was statistically significant (Estimate = 52.16, SE = 10.93, p < 0.001), indicating that, on average, each additional 1,000 tests is associated with approximately 52 more reported positive cases. The fixed intercept was not statistically significant (Estimate = 95.05, p = 0.281), suggesting no strong global baseline when adjusting for testing.

```{r}
#Generate model prediction
df_clean$predicted <- predict(model_refined)

#Plot all countries
ggplot(df_clean, aes(x = daily_tested_scaled, y = daily_positive, color = Country_Region)) + 
  geom_point(alpha = 0.3) +
  geom_line(aes(y = predicted, group = Country_Region), linewidth = 1) +
  labs(
    title = "Figure 1. LMM Predictions: All Countries",
    subtitle = "Fitted lines from Linear Mixed Model across all countries",
    x = "Daily Tests (in thousands)",
    y = "Daily Positive Cases"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```


### 3.5 Random Effects Interpretation
```{r}
VarCorr(model_refined)
```

The random intercept variance was 435,521 (SD = 659.94), indicating substantial variability in baseline positivity rates across countries. The random slope variance for daily_tested_scaled was 4,217 (SD = 64.94), suggesting that the effectiveness of testing (i.e., how many positives are identified per test) varies significantly by country. The negative correlation (-0.42) between random intercepts and slopes suggests that countries with higher baseline positivity tend to have a smaller marginal gain in detected positives per test.

### 3.6 Assumption Checks

The residuals vs. fitted plot revealed some mild heteroscedasticity, with greater variance in residuals at higher fitted values. However, no strong pattern of violation was observed. The Q-Q plot shows some deviation from normality at the tails, which is common in large public health datasets. Overall, the residuals appear reasonably well-behaved.

```{r}
# Select a subset of countries 
selected_countries <- c("United States", "India", "Brazil", "Germany", "South Korea")
df_subset <- df_clean[df_clean$Country_Region %in% selected_countries, ]
 
# Generate predictions for selected countries
df_subset$predicted <- predict(model_refined, newdata = df_subset)
 
# Plot
ggplot(df_subset, aes(x = daily_tested_scaled, y = daily_positive, color = Country_Region)) +
  geom_point(alpha = 0.6) +  # Raw data points
  geom_line(aes(y = predicted), linewidth = 1.2) +  # Use linewidth instead of size
  labs(
    title = "Figure 2: LMM Predictions for Selected Countries",
       subtitle = "Fitted lines for United States, India, Brazil, Germany, and South Korea",
   	x = "Daily Tests (in thousands)",
   	y = "Daily Positive Cases") +
  theme_minimal()

```
This scatterplot shows the relationship between testing and positive cases for specific countries and the country-wise differences in slopes (e.g., some countries have steeper slopes than others). In addition, this scatterplot shows how the model predicts positivity rates differently per country.

### 3.7 Visual Insights
Visualizations of fitted model predictions against observed data showed that:

- The relationship between testing and positivity varies across countries.

- Some countries exhibit steeper slopes (e.g., South Korea), while others show flatter or more variable trends (e.g., United States).

- This validates the use of a random slope model, as a single global slope would fail to capture such heterogeneity.



### 4 Conclusion

This study applied a Linear Mixed Model (LMM) to analyze COVID-19 testing data across countries using time-indexed, repeated measures. By modeling daily positive case counts as a function of daily testing volume and allowing both intercepts and slopes to vary by country, the LMM effectively captured differences in baseline positivity and testing effectiveness across global regions.

The fixed effect results demonstrated that increased testing was significantly associated with higher detected case counts, underscoring the critical role that testing plays in understanding and managing the spread of COVID-19. Additionally, the model’s random effects revealed meaningful variation between countries—both in baseline levels of positive cases and in how each country’s positivity rate responded to increased testing.

Model diagnostics showed acceptable residual behavior, though some mild heteroscedasticity and non-normality were present, likely due to the large scale and variability of the data. A minor convergence warning suggests future refinement of optimization settings may improve model stability, but the core results remain interpretable and valid.

Overall, the use of LMMs in this context illustrates their power in handling real-world public health data that is hierarchical, unbalanced, and repeated over time. The flexibility of LMMs allows researchers to extract more meaningful insights than would be possible with simpler linear models or ANOVA, particularly when assessing country-specific patterns during a global health crisis. Future research might explore more complex temporal structures or additional covariates (e.g., vaccination rates, policy interventions) to further enhance our understanding of COVID-19 dynamics worldwide.


## 5 References

