---
title: "Unlocking Sleep Health and Lifestyle insights using Normal Linear Mixed Models"
subtitle: "An analysis of normal linear mixed model algorithms and it's applications in the sleep world "
author: "Autumn Clark and Alex Ottley (Advisor: Dr.Cohen)"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

[Literature Review](litreview.html){target="_blank"} |
[Slides](slides.html){target="_blank"}

::: callout-important
**Remember:** Your goal is to make your audience understand and care
about your findings. By crafting a compelling story, you can effectively
communicate the value of your data science project.

Carefully read this template since it has instructions and tips to
writing!
:::

## Introduction

  In recent years, statistical models have become indispensable tools in understanding complex data structures across various fields, including biology, social sciences, economics, and medicine. One powerful class of models that has gained significant traction is the linear mixed model (LMM). Linear mixed models extend traditional linear regression by incorporating both fixed and random effects, making them well-suited for data with hierarchical or grouped structures. Unlike standard linear models, which assume that all observations are independent, linear mixed models allow for the modeling of correlation within clusters or repeated measures, thus providing more flexibility and accuracy in analyzing real-world data.

  LMMs are particularly useful in addressing challenges such as nested data structures, longitudinal measurements, and variability across subjects or experimental units. By allowing for random effects, such models account for unobserved heterogeneity, making them invaluable in situations where individuals or units differ in ways not captured by the fixed effects alone. This flexibility enhances the model's ability to generalize and improve predictions, making it a versatile tool in both theoretical and applied research.

  Normal Linear Mixed Models (LMMs) have become a fundamental tool in statistical modeling, particularly in hierarchical data structures. LMMs account for both fixed effects, which describe population-level influences and random effects. LMMs are different from other methods such as ANOVA. This flexibility makes them particularly valuable in medical and healthcare research, where patient responses vary.

  The following is hypothetical and will be modified as further direction for the project is taken such as datasets, methodology, design and analysis. The application of LMMs spans various domains, including health-related quality of life assessments, cost predictions, blood pressure monitoring and cardiovascular studies. This paper explores the role of normal linear mixed models in different medical and health research studies. This paper plans to examine advantages of LMMs over other methods such as ANOVA and linear regression. The goal of this paper is to take a data set preferably relating to healthcare and use LMMs to prove their efficiency for a particular data set such as blood pressure data. LMMs would be useful to analyze a blood pressure data set to model patient specific fluctuations. This paper will analyze a blood pressure data set using LMMs to capture circadian rhythms, REM sleep patterns and potential heart issues (hypothetical). 


## Methods

-   Detail the models or algorithms used.

-   Justify your choices based on the problem and data.

*The common non-parametric regression model is*
$Y_i = m(X_i) + \varepsilon_i$*, where* $Y_i$ *can be defined as the sum
of the regression function value* $m(x)$ *for* $X_i$*. Here* $m(x)$ *is
unknown and* $\varepsilon_i$ *some errors. With the help of this
definition, we can create the estimation for local averaging i.e.*
$m(x)$ *can be estimated with the product of* $Y_i$ *average and* $X_i$
*is near to* $x$*. In other words, this means that we are discovering
the line through the data points with the help of surrounding data
points. The estimation formula is printed below [@R-base]:*

$$
M_n(x) = \sum_{i=1}^{n} W_n (X_i) Y_i  \tag{1}
$$$W_n(x)$ *is the sum of weights that belongs to all real numbers.
Weights are positive numbers and small if* $X_i$ *is far from* $x$*.*

*Another equation:*

$$
y_i = \beta_0 + \beta_1 X_1 +\varepsilon_i
$$

## Analysis and Results

### Data Exploration and Visualization
Data Set: 

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

A study was conducted to determine how...

```{r, warning=FALSE, echo=T, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=TRUE}
# Load Data
kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

### Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

### Conclusion

-   Summarize your key findings.

-   Discuss the implications of your results.

## References
