---
title: "Gathering insights about COVID19 Worldwide Testing using Linear Mixed Models"
subtitle: "UWF IDC6940 Capstone Project on Linear Mixed Models (LMMs)"
author: "Alex Ottley and Autumn Clark (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
code_folding: true # or "hide" to collapse by default
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

More information about `revealjs`:
<https://quarto.org/docs/reference/formats/presentations/revealjs.html>
:::

## Introduction - Understanding LMMs {.smaller}

Linear Mixed Models (LMMs) are statistical models that extend
traditional linear regression by including both:

-   **Fixed effects:** These are the overall effects you're interested
    in studying (e.g., how testing volume affects case counts across all
    countries).

-   **Random effects:** These account for variation within subgroups or
    clusters (e.g., differences between countries), allowing intercepts
    and/or slopes to vary by group.

## Introduction - Understanding LMMs continued {.smaller}

LMMs are especially useful when:

-   Data is **hierarchical** or **nested** (like days within countries),

-   Measurements are repeated over time,

-   The data is **unbalanced** (not the same number of observations for
    each group).

## Methods - Software Packages used for LMMs {.smaller}

-   Linear mixed-effects models (LMMs) are widely used for analyzing
    data with both **fixed and random effects**, especially in fields
    involving **repeated measures** or **hierarchical data structures**.

-   A major advancement in the implementation of LMMs in R is the
    development of the **lme4** package, documented by Bates et
    al.(2015) [@BatesandDouglas2005]. This package provides a framework
    for specifying and fitting mixed-effects models, using sparse matrix
    methods and numerical optimization to handle complex random-effects
    structures.
```{r}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
library(lme4)
library(nlme)
library(dplyr)
library(ggplot2)
library(readr)
```
    

## Methods - Model Structure and Notation {.smaller}

The general form of a linear mixed model is: $$
y_{ij} = \beta_0 + \beta_1 x_{ij} + b_{0j} + b_{1j}x_{ij} + \varepsilon_{ij}
$$ Where,

-   $y_{ij}$: response variable for observation in $i$ in group $j$

-   $x_{ij}$: predictor (e.g., scaled daily tests)

-   $\beta_0,\beta_1$: fixed intercept and slope (population-level)

-   $b_{0j}, b_{1j}$: random intercept and slope for group $j$ (e.g.,
    country)

-   $\varepsilon_{ij}$: residual error, assumed to be normally
    distributed: $\varepsilon_{ij} \sim N (0, \sigma^2)$

## Methods - Model Structure and Notation continued {.smaller}

In matrix form, the LMM can be represented as:

$$
y = X\beta + Zb + \varepsilon
$$ Where,

-   $y$: vector of outcomes

-   $X$: fixed-effects design matrix

-   $\beta$: vector of fixed effect coefficients

-   $Z$: random-effects design matrix

-   $b$: vector of random effects, assumed $\sim N (0, G)$

-   $\varepsilon$: residual errors, assumed $\sim N(0, R)$

## Data Exploration and Visualization {.smaller}

-  Our data comes from a [Covid19 Worldwide Testing](https://www.kaggle.com/datasets/lin0li/covid19testing/data){target="_blank"} dataset on Kaggle. It includes daily test counts, daily positive cases, and other variables across different countries over time.

```{r echo=TRUE, warning=FALSE}
# Load Data
library(readr)
COVID19 <- read_csv("tested_worldwide.csv")
kable(head(COVID19))
df <- COVID19

```


## Exploratory Data Analysis {.smaller}
- Before modeling, we cleaned the data by removing rows with missing values in the daily test count, daily positive cases, and region columns.
- We alsoc scaled the test count variable by dividing it by 1,000 for easier interpretation.

```{r echo=TRUE, warning=FALSE}
df <- COVID19
#Convert Date to Date format
df$Date <- as.Date(df$Date)

#Remove rows with missing values in relevant columns
df_clean <- df %>%
  filter(!is.na(daily_positive), !is.na(daily_tested), !is.na(Country_Region))

#Rescale daily_tested column to improve interpretation and convergence
df_clean$daily_tested_scaled <- df_clean$daily_tested / 1000

```



## Data Exploration and Visualization {.smaller}

A study was conducted to determine how...


```{r, warning=FALSE, echo=F}

```

## Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

## Conclusion

-   Linear mixed models are powerful tools for analyzing complex,
    time-based public health data, capturing both global trends and
    country-specific differences.
-   This project helped us understand global COVID-19 testing patterns,     showing that the number of tests is a strong predictor of detected       cases, but the effect size varies by country.
- It also shows the value of LMMs over simpler models, like linear regression, when working with complex, international health data.


## References
