---
title: "Gathering insights about COVID19 Worldwide Testing using Linear Mixed Models"
subtitle: "UWF IDC6940 Capstone Project on Linear Mixed Models (LMMs)"
author: "Alex Ottley and Autumn Clark (Advisor: Dr. Cohen)"
date: '`r Sys.Date()`'
format:
  revealjs
course: Capstone Projects in Data Science
bibliography: references.bib # file contains bibtex for references
#always_allow_html: true # this allows to get PDF with HTML features
code_folding: true # or "hide" to collapse by default
self-contained: true
execute: 
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

More information about `revealjs`:
<https://quarto.org/docs/reference/formats/presentations/revealjs.html>
:::

## Introduction - Understanding LMMs  {.smaller}

Linear Mixed Models (LMMs) are statistical models that extend traditional linear regression by including both:

- Fixed effects: These are the overall effects you're interested in studying (e.g., how testing volume affects case counts across all countries).

- Random effects: These account for variation within subgroups or clusters (e.g., differences between countries), allowing intercepts and/or slopes to vary by group.


## Introduction - Understanding LMMs continued {.smaller}

LMMs are especially useful when:

- Data is hierarchical or nested (like days within countries),

- Measurements are repeated over time,

- The data is unbalanced (not the same number of observations for each group).




## Methods - Model Structure and Notation {.smaller}
The general form of a linear mixed model is:
$$
y_{ij} = \beta_0 + \beta_1 x_{ij} + b_{0j} + b_{1j}x_{ij} + \varepsilon_{ij}
$$
Where,

- $y_{ij}$: response variable for observation in $i$ in group $j$

- $x_{ij}$: predictor (e.g., scaled daily tests)

- $\beta_0,\beta_1$: fixed intercept and slope (population-level)

- $b_{0j}, b_{1j}$: random intercept and slope for group $j$ (e.g., country)

- $\varepsilon_{ij}$: residual error, assumed to be normally distributed: $\varepsilon_{ij} \sim N (0, \sigma^2)$

## Methods - Model Structure and Notation continued {.smaller}
In matrix form, the LMM can be represented as:

$$
y = X\beta + Zb + \varepsilon
$$
Where,

- $y$: vector of outcomes

- $X$: fixed-effects design matrix

- $\beta$: vector of fixed effect coefficients

- $Z$: random-effects design matrix

- $b$: vector of random effects, assumed $\sim N (0, G)$

- $\varepsilon$: residual errors, assumed $\sim N(0, R)$



## Data Exploration and Visualization

-   Describe your data sources and collection process.

-   Present initial findings and insights through visualizations.

-   Highlight unexpected patterns or anomalies.

## Data Exploration and Visualization {.smaller}

A study was conducted to determine how...

```{r, warning=FALSE, echo=F, message=FALSE}
# loading packages 
library(tidyverse)
library(knitr)
library(ggthemes)
library(ggrepel)
library(dslabs)
```

```{r, warning=FALSE, echo=F}
# Load Data
#kable(head(murders))

ggplot1 = murders %>% ggplot(mapping = aes(x=population/10^6, y=total)) 

  ggplot1 + geom_point(aes(col=region), size = 4) +
  geom_text_repel(aes(label=abb)) +
  scale_x_log10() +
  scale_y_log10() +
  geom_smooth(formula = "y~x", method=lm,se = F)+
  xlab("Populations in millions (log10 scale)") + 
  ylab("Total number of murders (log10 scale)") +
  ggtitle("US Gun Murders in 2010") +
  scale_color_discrete(name = "Region")+
      theme_bw()
  

```

## Modeling and Results

-   Explain your data preprocessing and cleaning steps.

-   Present your key findings in a clear and concise manner.

-   Use visuals to support your claims.

-   **Tell a story about what the data reveals.**

```{r}

```

## Conclusion

- Linear mixed models are powerful tools for analyzing complex, time-based public health data, capturing both global trends and country-specific differences.

_ 

## References
